---
title: "cca"
author: "Jordan"
date: "`r Sys.Date()`"
output: html_document
---


```{r}
# Load the required packages
library(dplyr)
library(vegan) # specpool () estimateR() poolaccum() estaccumR()
library(ggplot2)
library(forcats )
library("readxl") 
```

Importation des données.
```{r Import}
# Import
# Tour PC
setwd("D:/OneDrive - UMONS/UMONS/MA2/MEMOIRE/Statistique/R/STAT-MEMOIRE/RAW2")
# Laptop
#setwd("C:/Users/Jordan/OneDrive - UMONS/UMONS/MA2/MEMOIRE/Statistique/R/STAT-MEMOIRE/RAW2")

# Import
soil <- read_excel("Soil_R.xlsx")
SpecCondStat <- read_excel("SpecCondStat.xls")
```

```{r Nettoyage}
# Nom plus compact
SCS <- SpecCondStat

# Retirer le site de Condorcet
SCS <- filter(SCS, TOPO != "Condorcet" )

# Renommer plus simplement
rename(SCS, "sp" = "SPEC.TAXPRIO" ) -> SCS
rename(SCS, "sites" = "TOPO" ) -> SCS

# Retirer les observations contenant l'espèce : "XXX"
# SCS <- filter(SCS, SPEC.TAXPRIO != "Bombus (Bombus)  sp." )

# Remplacement des noms d'espèces désuets
SCS$sp[SCS$sp == "Bombus lucorum"] <- "Terrestribombus  sp."
SCS$sp[SCS$sp == "Bombus terrestris"] <- "Terrestribombus  sp."
SCS$sp[SCS$sp == "Chalicodoma ericetorum"] <- "Megachile ericetorum"
SCS$sp[SCS$sp == "Halictus tumulorum"] <- "Seladonia tumulorum"

# Remplacement du noms de sites
SCS$sites[SCS$sites == "Les Gourmandes de la Procession"] <- "Gourmandes de la Procession"
soil$Site[soil$Site == "Les Gourmandes de la Procession"] <- "Gourmandes de la Procession"

# retirer les sp abondantes (bombus)
# SCS <- filter(SCS, sp != "Bombus pascuorum")
# SCS <- filter(SCS, sp != "Bombus (Bombus)  sp.")
# SCS <- filter(SCS, sp != "Terrestribombus  sp.")
# SCS <- filter(SCS, sp != "Bombus vestalis")
# SCS <- filter(SCS, sp != "Bombus hypnorum")
# SCS <- filter(SCS, sp != "Bombus hortorum")
# SCS <- filter(SCS, sp != "Bombus pratorum")
# SCS <- filter(SCS, sp != "Bombus lapidarius")


#### Abbréger le nom des sites
# Liste des Potagers
Potagers <- c("Abbaye St-Denis", "Rue de l'Egalite", "Chasse Cambier", "Mel Legumes", "Gourmandes de la Procession", "Jean d'Avesnes", "Parc du bois de Mons", "Jardin Suspendu", "Ecole de l'Esperance", "Fond du petit marais")
# Liste des Parcs
Parcs <- c("Village des abeilles", "Siege social", "Parc du Beffroi", "Stievenart", "Parc Bonaert")

### Sol
## Metaux
# Créer une colonne Code en fonction des types de sites
soil$Code <- fct_collapse(
    soil$Site,
    "Potager" = Potagers,
    "Parc" = Parcs)

# Garder pour rownames
soil_site_name <- select(soil, Site)

# Garder les métaux seulement
metal <- soil[, c("As", "Cd", "Cr", "Cu", "Hg", "Pb", "Ni", "Zn")]
row.names(metal) <- soil_site_name$Site


```

```{r }
# Preparation des données 
data_bee_sp <- select(SCS, sp, N, sites)
data_bee_sp <- as.data.frame(data_bee_sp)
data_bee_sp <- aggregate(N ~ sp + sites, data = data_bee_sp, sum)
data_bee_sp <- xtabs(N ~ sites + sp, data_bee_sp)
data_bee_sp <- type.convert(data_bee_sp)
```

```{r}
str(metal)
summary(metal)
```





Le pourcentage de variation expliqué par les axes est trop faible lorsque l'on utilise tous les axes.
```{r}
spe <- data_bee_sp
# le "." signifie contre toutes les var env; 1 : signifie aucune var env
spe.cca <- cca(spe ~ ., metal)
summary(spe.cca) # Scaling 2 (default)
# Unadjusted and adjusted R^2 - like statistics
RsquareAdj(spe.cca)
```

```{r}
# cca(formula = spe ~ Argile + `Limon fin` + `Limon grossier` + `Sable fin` + `Sable grossier`, data = as.data.frame(Granulo_scale))
# test l'ensemble de la CCA si le modèle explique une quantité significatif de la variation dans les données
anova(spe.cca) 
# test pour chacune des var env
anova(spe.cca, by = "term", permutations = 999)
```


```{r}
vif.cca(spe.cca)
```

En supprimant les valeurs supérieurs à 10, on obtient 
```{r}
spe <- data_bee_sp
# c( "As", "Cr","Cd", "Cu", "Hg", "Pb","Ni", "Zn")
spe.cca <- cca(spe ~ ., metal[c( "Ni","As", "Cr", "Cd", "Hg", "Pb")])
#summary(spe.cca) # Scaling 2 (default)
# Unadjusted and adjusted R^2 - like statistics
# RsquareAdj(spe.cca)
# cca(formula = spe ~ Argile + `Limon fin` + `Limon grossier` + `Sable fin` + `Sable grossier`, data = as.data.frame(Granulo_scale))
# test l'ensemble de la CCA si le modèle explique une quantité significatif de la variation dans les données
anova(spe.cca) 
# test pour chacune des var env
anova(spe.cca, by = "term", permutations = 999)
```

```{r}
# cca(formula = spe ~ Argile + `Limon fin` + `Limon grossier` + `Sable fin` + `Sable grossier`, data = as.data.frame(Granulo_scale))
# test l'ensemble de la CCA si le modèle explique une quantité significatif de la variation dans les données
anova(spe.cca) 
# test pour chacune des var env
anova(spe.cca, by = "term", permutations = 999)
```

```{r}
#vegan::plot.cca(spe.cca)
plot(spe.cca, scaling = 3, display ="wa")
plot(spe.cca, scaling = 3, display ="lc")
plot(spe.cca, scaling = 3, display ="bp")
plot(spe.cca, scaling = 3, display ="reg")
plot(spe.cca, scaling = 3, display ="cn")
plot(spe.cca, scaling = 3, display ="all")
```
```{r}
plot(spe.cca, scaling = 0, display ="all") # score brute non mise à l'échelle
plot(spe.cca, scaling = 1, display ="all") # site : eigenvelues mise à l'échelle en fonction des sites 
plot(spe.cca, scaling = 2, display ="all") # espèce : eigenvelues mise à l'échelle en fonction des espèces 
plot(spe.cca, scaling = 3, display ="all") # symétrique : eigenvelues mise à l'échelle en fonction des espèces et sites (racine carré des eigenvalues)
```

```{r}
# CCA scaling 1 biplot without species (using lc site scores)
plot(spe.cca, scaling = 1, display = c("lc", "cn"), main = "Biplot CCA spe ~ env3 - scaling 1")
```

```{r}
par(mfrow = c(1, 2))
# Scaling 1: species scoresscaled to the relative eigenvalues,
# sites are weighted averages of the species
plot(spe.cca,
scaling = 1,
display = c("sp", "lc", "cn"),
main = "Triplot CCA spe ~ env3 - scaling 1"
)
# Default scaling 2: site scores scaled to the relative
# eigenvalues, species are weighted averages of the sites
plot(spe.cca,
display = c("sp", "lc", "cn"),
main = "Triplot CCA spe ~ env3 - scaling 2"
)

```

```{r}
# CCA scaling 1 biplot without species (using lc site scores)
plot(spe.cca,
scaling = 1,
display = c("lc", "cn"),
main = "Biplot CCA spe ~ env3 - scaling 1"
)
```



```{r}
# CCA scaling 2 biplot with species but without sites
plot(spe.cca,
scaling = 2,
display = c("sp", "cn"),
main = "Biplot CCA spe ~ env3 - scaling 2"
)
```

Permutation Tests in CCA,Forward Selection
and Parsimonious CCA
CCA results can be tested for significance by permutation, in the same way as RDA.
```{r}
# Permutation test of the overall analysis
anova(spe.cca, permutations = how(nperm = 999))
# Permutation test of each axis
anova(spe.cca, by = "axis", permutations = how(nperm = 999))

```

```{r}
# CCA-based forward selection using vegan's ordistep()
# This function allows the use of factors like 'slo' in env3
# cca.step.forward <-
# ordistep(cca(spe ~ 1, data = env3),
# scope = formula(spe.cca),
# direction = "forward",
# permutations = how(nperm = 199))
```


```{r}
# calcul du VIF (corrélation entre variable)
#vif <- diag(solve(cor(X)))

# Variance inflation factors (VIF) in two RDAs
# First RDA of this Chapter: all environmental variables
# except dfs
vif.cca(spe.cca)
# Partial RDA – physiographic variables only
#vif.cca(spechem.physio)

```

```{r}
# CCA-based forward selection using vegan's ordistep()
# This function allows the use of factors like 'slo' in env3
# cca.step.forward <-
# ordistep(cca(spe ~ 1, data = env3),
# scope = formula(spe.cca),
# direction = "forward",
# permutations = how(nperm = 199))
```



```{r}
## Parsimonious CCA using ele, oxy and bod
# spe.cca.pars <- cca(spe ~  oxy + bdo, data = env3)
# anova(spe.cca.pars, permutations = how(nperm = 999))
# anova(spe.cca.pars, permutations = how(nperm = 999), by = "axis")
# # R-square – like statistics
# RsquareAdj(spe.cca.pars)
# # Compare variance inflation factors
# vif.cca(spe.cca)
# vif.cca(spe.cca.pars)
```
Three-Dimensional Interactive Plots

```{r}
#library(vegan3d)
# Plot of the sites only (wa scores)
# ordirgl(spe.cca.pars, type = "t", scaling = 1)
# # Connect weighted average scores to linear combination scores
# orglspider(spe.cca.pars, scaling = 1, col = "purple")
```

```{r}
# Plot the sites (wa scores) with a clustering result
# Colour sites according to cluster membership
# gr <- cutree(hclust(vegdist(spe.hel, "euc"), "ward.D2"), 4)
# ordirgl(spe.cca.pars,
# type = "t",
# scaling = 1,
# ax.col = "black",
# col = gr + 1
# )
# # Connect sites to cluster centroids
# orglspider(spe.cca.pars, gr, scaling = 1)
```


```{r}
# Complete CCA 3D triplot
# ordirgl(spe.cca.pars, type = "t", scaling = 2)
# orgltext(spe.cca.pars,
# display = "species",
# type = "t",
# scaling = 2,
# col = "cyan"
# )
# # Plot species groups (Jaccard dissimilarity, useable in R mode)
# gs <-
# cutree(
# hclust(vegdist(t(spe), method = "jaccard"), "ward.D2"),
# k = 4)
# ordirgl(spe.cca.pars,
# display = "species",
# type = "t",
# col = gs + 1)
```



```{r}
ca(data_bee_sp)
summary(ca(data_bee_sp))
cca(data_bee_sp)
summary(cca(data_bee_sp))
```



```{r}
# https://stackoverflow.com/questions/58957101/how-do-i-customize-ordination-plot-in-vegan

varechem <- as.data.frame(metal[c( "As", "Cr","Cd", "Hg", "Pb","Ni")])
varespec <- as.data.frame(data_bee_sp)
#CCA
cca_model<-cca(varespec ~ .,data=varechem)
plot(cca_model,choices=c(1,2), display=c('sp','bp'), scaling=2)

#Get CCA scores
df_species  <- data.frame(summary(cca_model)$species[,1:2])# get the species CC1 and CC2 scores
df_species  <- as.data.frame(df_species) 
df_environ  <- scores(cca_model, display = 'bp') #get the environment vars CC1 and CC2 scores
df_environ  <- as.data.frame(df_environ)   

cca1_varex<-round(summary(cca_model)$cont$importance[2,1]*100,2) #Get percentage of variance explained by first axis
cca2_varex<-round(summary(cca_model)$cont$importance[2,2]*100,2) #Get percentage of variance explained by second axis

#Set a scaling variable to multiply the CCA values, in order to get a very similar plot to the the one generated by plot(cca_model). You can adjust it according to your data
scaling_factor <- 2

ggplot(df_species, 
       aes(x=CCA1, y=CCA2)) + 
  #Draw lines on x = 0 and y = 0
  geom_hline(yintercept=0, 
             linetype="dashed") +
  geom_vline(xintercept=0, 
             linetype="dashed") +
  coord_fixed()+
  #Add species text
  geom_text(data=df_species, 
            aes(x=CCA1,#Score in CCA1 to add species text
                y=CCA2,#Score in CCA2 to add species text
                label=rownames(df_species),
                hjust=0.5*(1-sign(CCA1)),#Set the text horizontal alignment according to its position in the CCA plot
                vjust=0.5*(1-sign(CCA2))),#Set the text vertical alignment according to its position in the CCA plot
            color = "forestgreen")+
  #Add environmental vars arrows
  geom_segment(data=df_environ, 
               aes(x=0, #Starting coordinate in CCA1 = 0 
                   xend=CCA1*scaling_factor,#Ending coordinate in CCA1  
                   y=0, #Start in CCA2 = 0
                   yend=CCA2*scaling_factor), #Ending coordinate in CCA2 
               color="firebrick1", #set color
               arrow=arrow(length=unit(0.01,"npc"))#Set the size of the lines that form the tip of the arrow
               )+
  #Add environmental vars text
  geom_text(data=df_environ, 
            aes(x=CCA1*scaling_factor, 
                y=CCA2*scaling_factor,
                label=rownames(df_environ),
                hjust=0.5*(1-sign(CCA1)),#Add the text of each environmental var at the end of the arrow
                vjust=0.5*(1-sign(CCA2))),#Add the text of each environmental var at the end of the arrow 
            color="firebrick1")+
  #Set bw theme
  theme_bw()+
  #Set x and y axis titles
  labs(x=paste0("CCA1 (",cca1_varex," %)"),
       y=paste0("CCA2 (",cca2_varex," %)"))
```

```{r}
# https://blogs.ncl.ac.uk/mep/2018/04/08/reproducible-publication-quality-multivariate-plots-in-r/
library(ggrepel)

varechem <- as.data.frame(metal[c( "As", "Cr","Cd", "Hg", "Pb","Ni")])
varespec <- as.data.frame(data_bee_sp)
vare_cca <- cca(spe ~ ., metal[c( "As", "Cr","Cd", "Hg", "Pb","Ni")])


vare_spp_sco <- scores(vare_cca, display = "species")
vare_sam_sco <- scores(vare_cca, display = "sites")
vare_env_sco <- scores(vare_cca, display = "bp")
vare_spp_tbl <- as_tibble(vare_spp_sco)
vare_sam_tbl <- as_tibble(vare_sam_sco)
vare_env_tbl <- as_tibble(vare_env_sco)
vare_spp_tbl <- mutate(vare_spp_tbl, vgntxt=rownames(vare_spp_sco),
                       ccatype = "species")
vare_sam_tbl <- mutate(vare_sam_tbl, vgntxt=rownames(vare_sam_sco),
                       ccatype = "sites")
vare_env_tbl <- mutate(vare_env_tbl, vgntxt=rownames(vare_env_sco),
                       ccatype = "bp")

ggplot(vare_spp_tbl, aes(x = CCA1, y = CCA2, label = vgntxt)) +
       geom_point() +
       geom_text_repel(seed = 123)



rescaled <- vare_env_tbl %>% 
            select(CCA1, CCA2) %>%
            as.matrix() * 1.5
vare_tbl <- select(vare_env_tbl, vgntxt, ccatype) %>%
            bind_cols(as_tibble(rescaled)) %>%
            bind_rows(vare_spp_tbl)

ggplot() +
  geom_point(aes(x=CCA1, y=CCA2), data=filter(vare_tbl, ccatype=="species"))  +
  geom_text_repel(aes(x=CCA1, y=CCA2, label=vgntxt, size=3.5),
                  data=vare_tbl, seed=123) + 
  geom_segment(aes(x=0, y=0, xend=CCA1, yend=CCA2), arrow=arrow(length = unit(0.2,"cm")),
               data=filter(vare_tbl, ccatype=="bp"), color="blue") +
  coord_fixed() +
  theme_bw() +
  theme(legend.position="none")
```

```{r}
# cca(formula = spe ~ Argile + `Limon fin` + `Limon grossier` + `Sable fin` + `Sable grossier`, data = as.data.frame(Granulo_scale))
# test l'ensemble de la CCA si le modèle explique une quantité significatif de la variation dans les données
anova(vare_cca) 
# test pour chacune des var env
anova(vare_cca, by = "term", permutations = 999)

RsquareAdj(vare_cca)
```

```{r}
permutest(vare_cca, permutations = 999)
```

```{r}
permutest(vare_cca, by = "term", permutations = 999)
```

```{r}
vif.cca(vare_cca)
```


```{r}
summary(vare_cca)
```

# Espèce terricole

```{r Import, message=FALSE, warning=FALSE, include=FALSE, results='hide', paged.print=FALSE}
# Tour PC
setwd("D:/OneDrive - UMONS/UMONS/MA2/MEMOIRE/Statistique/R/STAT-MEMOIRE/RAW2/Traits abeilles")
# Laptop
#setwd("C:/Users/Jordan/OneDrive - UMONS/UMONS/MA2/MEMOIRE/Statistique/R/STAT-MEMOIRE/RAW2/Traits abeilles")

# Import
traits <- read_excel("Trait-bee-each-site.xlsx")

# Réécriture des noms de familles
traits$familly[traits$familly == "ANDRENIDAE"] <- "Andrenidae"
traits$familly[traits$familly == "APIDAE"] <- "Apidae"
traits$familly[traits$familly == "COLLETIDAE"] <- "Colletidae"
traits$familly[traits$familly == "HALICTIDAE"] <- "Halictidae"
traits$familly[traits$familly == "MEGACHILIDAE"] <- "Megachilidae"
traits$familly[traits$familly == "MELITTIDAE"] <- "Melittidae"

# Remplacement du noms de sites
traits$sites[traits$site == "Les Gourmandes de la Procession"] <- "Gourmandes de la Procession"
```

```{r}
terricole <- filter(traits, Nest_area == "Soil")

# Preparation des données 
terricole <- select(terricole, sp, N, site)
terricole <- as.data.frame(terricole)
terricole <- aggregate(N ~ sp + site, data = terricole, sum)
terricole <- xtabs(N ~ site + sp, terricole)
terricole <- type.convert(terricole)
```

```{r}
varechem <- as.data.frame(metal[c( "As", "Cr","Cd", "Hg", "Pb","Ni")])
varespec <- as.data.frame(terricole)
vare_cca <- cca(spe ~ ., metal[c( "As", "Cr","Cd", "Hg", "Pb","Ni")])
```

```{r}
# cca(formula = spe ~ Argile + `Limon fin` + `Limon grossier` + `Sable fin` + `Sable grossier`, data = as.data.frame(Granulo_scale))
# test l'ensemble de la CCA si le modèle explique une quantité significatif de la variation dans les données
anova(vare_cca) 
# test pour chacune des var env
anova(vare_cca, by = "term", permutations = 999)

RsquareAdj(vare_cca)
```
